Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-20 index.qmd

The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing. Thus, a reader looking at published code can often be left with many questions about why certain choices were made. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

@@

To quantify redundancy, the concept of mutual information is used. Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the KL-distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:

@@

In order to test our development of analysis validation checks based on the New York City data, we incorporate data from 39 additional cities from the original NMMAPS study and apply our checks to those data. These cities (along with New York City) represent approximately the largest 40 metropolitan areas in the United States. The complete list of cities used are presented in @tbl-all-cities. For each city, we had daily data on all-cause mortality, PM10, and average temperature for the years 1992--2000.

```{r}
#| label: tbl-all-cities

@@

df |>
  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
                ~ifelse(.x, "T", "F"))) |>
  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |> 
  kable(format = "latex", booktabs = TRUE, escape = F, 
        align = "lrccrrcc",
        col.names = linebreak(c("City", "PM10\nCoefficient", "Outcome\nUnexpected", "Predicted\nUnexpected", 
                                "Correlation\nMortality\nPM10", "Correlation\nMortality\nTemperature",
                                "Mortality\nOutlier", "PM10\nOutlier"))) |>
  kableExtra::kable_styling(latex_options="scale_down") 
```

We applied each of the analysis validation checks and the fitted logic regression model separately to each of the cities' data to make predictions of whether the outcome will be unexpected or not. We also applied the generalized linear model to the data from each of the cities to estimate the association between PM10 and mortality in each city. The estimates of the association and the results of each of the analysis validation checks for each city are shown in @tbl-all-cities.

```{r}
#| label: tbl-accuracy
df |> 
  dplyr::select(unexpect, .pred) |> 
  table() |> 
  as_tibble() |> 
  rename(Actual = unexpect) |> 
  pivot_wider(names_from = .pred, values_from = n) |> 
  kable(caption = "A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.", 
        format = "latex", booktabs = TRUE) |> 
   add_header_above(c("", "Prediction"=2), escape = FALSE) 
```

@tbl-accuracy summaries the prediction accuracy of the logic regression model for the 40 cities. The model correctly predicts `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy of `r label_percent()(accurate/nrow(df))`. Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r df |> filter(unexpect, .pred) |> pull(cityname)`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)`. These cities are flagged due to failure in both mortality-PM10 correlation and mortality-temperature correlation checks. Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were a false positive, resulting in a precision of `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`.

@@

@@

Systematically generating realistic simulated data is a key component of our approach and is deserving of further consideration. In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature. However, simulation can become complex when additional restrictions are imposed or when a greater range of scenarios is desired. In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.

>>>>

The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing. Thus, a reader looking at published code can often be left with many questions about why certain choices were made. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

@@

To quantify redundancy, the concept of mutual information is used. Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the Kullback-Leibler distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:

@@

In order to test our development of analysis validation checks based on the New York City data, we incorporate data from 39 additional cities from the original NMMAPS study and apply our checks to those data. These cities (along with New York City) represent approximately the largest 40 metropolitan areas in the United States by population. The complete list of cities used are presented in @tbl-all-cities. For each city, we had daily data on all-cause mortality, PM10, and average temperature for the years 1992--2000.

```{r}
#| label: tbl-all-cities
#| tbl-cap: Results from full analysis of 40 cities from the NMMAPS air pollution and mortality study. In the table, a "T" indicates that a check failed and an "F" indicates that a check passed.

@@

df |>
  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
                ~ifelse(.x, "T", "F"))) |>
  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |> 
  kable(format = "latex", booktabs = TRUE, escape = F, 
        align = "lrccrrcc",
        col.names = linebreak(c("City", 
                                "PM10\nEstimated\nCoefficient",
                                "Outcome\nObserved\nUnexpected", 
                                "Logic Regression\nPredicted\nUnexpected", 
                                "Correlation:\nMortality-\nTemperature",
                                "Correlation:\nMortality-\nPM10",
                                "Mortality\nOutlier", 
                                "PM10\nOutlier"))) |>
  kableExtra::kable_styling(latex_options="scale_down") 
```

We applied each of the analysis validation checks and the fitted logic regression model separately to each of the cities' data to make predictions of whether the outcome will be unexpected or not. We also applied the generalized linear model to the data from each of the cities to estimate the association between PM10 and mortality in each city, adjusted for temperature. The estimates of the association and the results of each of the analysis validation checks for each city are shown in @tbl-all-cities.

```{r}
#| label: tbl-accuracy
#| tbl-cap: Summary of the observed and the predicted unexpected PM10 coefficient results from the 40 NMMAPS cities using the logic regression model.

df |> 
  dplyr::select(unexpect, .pred) |> 
  table() |> 
  as_tibble() |> 
  rename(Observed = unexpect) |> 
  pivot_wider(names_from = .pred, values_from = n) |> 
  kable(format = "latex", 
        booktabs = TRUE) |> 
   add_header_above(c("", "Predicted"=2), 
                    escape = FALSE) 
```

```{r,include=FALSE}
cities_accurate <- df |> 
  filter(unexpect, .pred) |> 
  pull(cityname)
cities_missed <- df |> 
  filter(unexpect, !.pred) |> 
  pull(cityname)
```

It is clear from @tbl-all-cities that some of the estimated PM10 coefficients are outside the expected range of [0, 0.005]. Eight of the 40 cities had negative estimated coefficients and were therefore considered unexpected by our original criterion. (It is perhaps worth noting that none of the unexpected outcomes was in the positive direction.) @tbl-accuracy summarizes the prediction accuracy of the logic regression model for the 40 cities. The model correctly predicts the PM10 coefficient status of `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy rate of `r label_percent()(accurate/nrow(df))`. Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model). These cities were flagged due to failures in both the mortality-PM10 correlation and mortality-temperature correlation analysis checks. Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were false positives, resulting in a precision of only `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`. This precision value is substantially lower than what was estimated by the simulation procedure (see @tbl-linear-reg) and suggests that the simulation does not adequately capture some features of the data generation process.

@@

In @sec-pm10-mortality we used the analysis checks in a diagnostic manner to examine the eight cities whose PM10 coefficients were unexpected. There, we found that seven of the cities had a mortality-PM10 correlation that was more negative than expected while also having a mortality-temperature correlation that was somewhat larger (less negative) than expected. What this result implies for the broader analysis depends on a number of factors, but the interplay between mortality, PM10, and temperature may warrant further investigation [see e.g. @welty2005acute]. Note that simply because the PM10 coefficient estimates were unexpected by our criterion, we do not mean to imply that they are ``wrong" in any sense. Indeed, negative estimated coefficients have been found in other studies of this nature [@bell2004ozone]. Rather, in this type of analysis, it may be that the interval for expected results needs to be revised. Further, it may be necessary to incorporate other outputs from the analysis, such as measures of uncertainty.

@@

Systematically generating realistic simulated data is a key component of our approach and is deserving of further consideration. In the PM10 example in @sec-pm10-mortality, the inverse transform method was used to preserve the correlation structure among mortality, PM10, and temperature. However, the simulation process can become complex when additional restrictions are imposed or when a greater range of scenarios is desired. In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data. The results in @sec-pm10-mortality on the broader NMMAPS dataset suggest that our simulation procedure may have been inadequate in reflecting the range of possible configurations that the data could take. In particular, our split-data approach, using New York City to guide the simulation and then applying the analysis checks to other cities, may not have been ideal. It may be worth exploring some recent work in data thinning [@neufeld2024data] and data fission [@leiner2023data] to create training datasets that are more representative of future data.
